BOLD Parameter Estimation Using Particle Filters

### 2010-07-15 doc/thesis/results.tex
updated some discussion about the first low noise results

\chapter{Simulation}

\section{Single Timeseries}
I performed two types of simulations. First, I simulated a single BOLD time-series based
on a randomly selected set of model parameters. Second I used a modified version of the
FSL tool possum to generate a noisy FMRI image.

This process was relatively straight forward,
given the state-space equations for the BOLD signal. After a "true" signal was generated,
I then added a identically and independently distributed (I.I.D.) Gaussian noise and a Wiener
process with Gaussian I.I.D. steps. Finally a carrier level was added, since BOLD is typically
measured as a \% difference from the
base level. Adding a carrier level meant that the exact same algorithm which was used for 
real data could be used for the simulated data.

Once this noisy simulated time series was generated, the particle filter algorithm
 was run on this single timeseries image. In order to find the best setting for the
particle filter I ran quite a few tests. Here I will include two sets of tests 
to determine the power of the particle filter in modeling. The first test is
low noise $\{\sigma_y = .001, \sigma_x = .0005\}$, while the second set of
tests had noise of $ \{\sigma_y = .01, \sigma_x = .005\} $, where $\sigma_y$ is the
measurement noise, and $\sigma_x$ is the wiener step size. Both signals used the
parameters ($\tau_0 = 1.45, \alpha = .3, E_0 = .47, V_0 = .044, \tau_s = 1.94, \tau_f = 1.99, \epsilon = 1.8$).
The particle filter used the parameters defined in \autoref{tab:Prior} (\autoref{sec:PriorDistrib}),
thus the particle filter was not centered over the correct values.

\subsection{Low Noise}
For the low noise case, the ten realization are shown in \autoref{fig:LowNoiseRealization}.
\begin{figure}
\label{fig:LowNoiseRealization}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/realization_lownoise}
\caption{Test Signals with low noise compared to the clean signal}
\end{figure}

The resulting fits compared to the \emph{true} BOLD signal are shown in \autoref{fig:FitComparisonLowNoise}.
The fits are good, but the noise is relatively low. The preprocessed signal compared
to the true BOLD signal is shown in \autoref{fig:PreprocessedLowNoise}.

\begin{figure}
\label{fig:PreprocessedLowNoise}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/preprocessed_lownoise}
\caption{A comparison of the preprocessed signals for the low noise case.}
\end{figure}

The preprocessing consists of several steps, which are discussed in detail in \autoref{sec:Preprocessing}.
Obviously the spline struggles a little bit at the end, as the graph shows, although 
overall the fit is pretty well. Finally, the set of fits to the preprocessed data, and in 
essence to the "true" BOLD signal is shown in \autoref{fig:FitComparisonLowNoise}.

\begin{figure}
\label{fig:FitComparisonLowNoise}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/comparison_lownoise}
\caption{A comparison of the fitted signals for the low noise case.}
\end{figure}

It becomes clear now the reason for the term particle filter. In several locations the output
of the particle filter looks like a filtered version of the input. For instance toward the
end, the estimates stay flat in spite of the preprocessed data drifting off a bit. By
this point, the algorithm had converged sufficiently to know that that wasn't possible.
A similar circumstance occurs at 100 seconds in. In general the fits are good, although because
of the slightly higher signal levels in the preprocessed signals, the fits tend to overestimate
the peak a bit. The final parameter sets are shown in \autoref{tab:LowNoiseParams}
\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c | c |}
\hline 
$\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$  & $ \sum \tau $ & $\sqrt{MSE}$ (Res.) &$\sqrt{MSE}$\\
\hline 
\rowcolor[gray]{.8}
1.45 & .3 & .47 & .044 & 1.94 & 1.99 & 1.8  & 5.38 &  & \\
\hline 
\hline 
1.22214 & 0.3449 & 0.33462 & 0.07138 & 1.60446 & 2.27529 & 1.59454 & 5.10189  &  0.00321067  & 0.00987647  \\
1.37493 & 0.33183 & 0.36296 & 0.07327 & 1.64076 & 2.10299 & 1.57627 & 5.11867 &  0.00305475  & 0.00993227  \\
1.16604 & 0.32205 & 0.34057 & 0.08217 & 1.64768 & 2.35351 & 1.24515 & 5.16722 &  0.00328932  & 0.00967961  \\
1.23176 & 0.32707 & 0.34028 & 0.07961 & 1.62698 & 2.18519 & 1.30326 & 5.04394 &  0.00284719  & 0.00912005  \\
1.1832 & 0.31787 & 0.34718 & 0.0821 & 1.54961 & 2.29115 & 1.27817 & 5.02396   &  0.00300634  & 0.00971258  \\
1.1424 & 0.33395 & 0.34725 & 0.07366 & 1.62208 & 2.29084 & 1.4025 & 5.05531   &  0.00283287  & 0.00948483  \\
1.30041 & 0.3596 & 0.35643 & 0.07679 & 1.56406 & 2.13234 & 1.60338 & 4.99681  &  0.00302802  & 0.01021885  \\
1.24008 & 0.34601 & 0.33978 & 0.08903 & 1.64989 & 2.23655 & 1.29004 & 5.12651 &  0.00304378  & 0.01007964  \\
1.1709 & 0.32739 & 0.34644 & 0.08255 & 1.53734 & 2.28262 & 1.37828 & 4.99087  &  0.00334488  & 0.01032886  \\
1.18967 & 0.34344 & 0.33554 & 0.07976 & 1.53582 & 2.30746 & 1.42774 & 5.03295 &  0.00317542  & 0.01001503  \\
1.184 & 0.34053 & 0.35017 & 0.08917 & 1.61025 & 2.27926 & 1.16448 & 5.07352   &  0.00288855  & 0.00950536  \\
\hline                                                                           
1.21868 & 0.33588 & 0.34557 & 0.07995 & 1.59899 & 2.24884 & 1.38762 & 5.06651 & 0.00306562     & 0.00981396 \\
\hline 
\end{tabular}
\caption{Estimated Parameters on 11 different runs with low noise. First row is the true parameters,
last is mean over the 10 runs. The $\sqrt{MSE}$ (Res.) is the square root of the mean square of the
residuals. Essentially this is the is the MSE between the estimated signal and the noisy signal which 
was available to the particle filter. Square root of MSE is the actual \emph{error}, with the true signal,
which of course was not available to the particle filter.}
\label{tab:LowNoiseResults} 
\end{table}

There are a few things worth noting here. First the time constants vary greatly across
runs with different noise realizations, yet the sum of the individual time constants
, $\tau_f$, $\tau_s$ and $\tau_0$ seem to be more consistent. In general the 
time constants are falling short of the true time constant. This could be a limitation
based on the prior distribution (which notably has mean values below the values used
in the simulation) or it could be caused by the delayed benefit of having a 
correct time constant. Even if one particle has a better time constant than another, if
the difference isn't severe, by the time this makes a difference in the weight, the 
particle with the better $\tau$ will have been weighted based on other parameters several
times. Another interesting result in the huge variation in the levels of $V_0$. In general,
with this admittedly small amount of noise, it would appear that the relation between a set
of parameters/stimuli with the output is not injective. In other words, it would appear
that a timeseries is not unique to a set of parameters. This is good justification that 
simultaneous blood volume or tagged flow calculations with the conventional FMRI 
could definitely benefit the model.

\begin{table}[t]
\centering
\begin{tabular}{|c | c  c  c  c  c  c  c |}
\hline 
  & $\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$ \\
\hline 
\rowcolor[gray]{.8} $\tau_0$ &   0.0189481 & -0.0014269 & -0.0011267 & -1.13e-05 & -0.0025616 & -0.0189559 & 0.0070405 \\
$\alpha$ &                       -0.0014269 & 0.0026716 & 9.93e-05 & -0.0002041 & -0.0008632 & -0.0016823 & 0.0071891 \\
\rowcolor[gray]{.8} $E_0$    &   -0.0011267 & 9.93e-05 & 0.0010701 & -0.0002277 & -0.0001177 & 0.0001013 & 0.0016972 \\
$V_0$    &                       -1.13e-05 & -0.0002041 & -0.0002277 & 0.0005401 & 4.3e-06 & 4.56e-05 & -0.0080494 \\
\rowcolor[gray]{.8} $\tau_s$ &   -0.0025616 & -0.0008632 & -0.0001177 & 4.3e-06 & 0.0128056 & 0.012878 & -0.005516 \\
$\tau_f$ &                       -0.0189559 & -0.0016823 & 0.0001013 & 4.56e-05 & 0.012878 & 0.0416927 & -0.0158182 \\
\rowcolor[gray]{.8} $\epsilon$&  0.0070405 & 0.0071891 & 0.0016972 & -0.0080494 & -0.005516 & -0.0158182 & 0.1567165 \\
\hline 
\end{tabular}
\caption{Typical Covariance matrix of the parameters at the end of a run.}
\label{tab:CovSim} 
\end{table}

@@

\begin{figure}
\subfigure[Converging histogram for $Q$, $S$, $F$, and $BOLD$ of the first run, low noise simulation.]
{\includegraphics[trim=7cm 2cm 7cm 2cm, width=15cm]{images/converge_lownoise3}}
\label{fig:LowNoiseHist}
\end{figure}
The covariance matrix (\autoref{tab:CovSim}) also confirms the idea that the $\tau$ parameters are interchangeable
as far as the BOLD signal goes. Notice the covariance of $\tau_f$ and $\tau_0$ is $-0.019$ whereas
the variance of $\tau_0$ and $\tau_f$ are $0.019$ and $0.04$ respectively. Clearly they are
related. The convergence properties of the first run in \autoref{tab:LowNoiseResults} may be
viewed in \autoref{fig:LowNoiseHist}.

>>>>

\chapter{Simulation}
Two levels are discussed in this section. First the simulation of single time series.
This is necessary to investigate the power of particle filters in 
identifying the BOLD model in a noisy environment. If the particle filter is unable
to get back to the "true" time-series in spite of noise then the particle filter
will not be useful. Additionally, determination of reasonable parameters from the
time-series is also desirable. The second simulation technique will be on a 
large group of voxels. Although the current technique treats each voxel separately,
the "Possum" FMRI simulation tool models noise more realistically, and provides additional
incite into the applicability of particle filters in modeling the BOLD signal on a large
scale.

\section{Single Timeseries}
Given the state-space equations for the BOLD signal, simulating a single time 
series is relatively straight forward. After a "true" signal is generated,
I then add identically and independently distributed (I.I.D.) Gaussian noise and a Wiener
process with Gaussian, I.I.D. steps. Finally a "carrier level" is added, since BOLD is typically
measured as a \% difference from the baseline. Although technically the particle filter
algorithm will immediately remove this, and calculate the \% difference, adding a carrier 
level means that the exact same algorithm which will is used for 
simulated data may be used for the real data.

Once this noisy simulated time series has been generated, the particle filter algorithm
 is then run on this single time series "image". Here I will include two sets of tests 
to determine the power of the particle filter in modeling. The first set are 
used to demonstrate the ability of the particle filter to find the most likely
set of parameters/state variables over the course of the run. The second type
of test will investigate the problem of false-positives. As the first
round of tests will show, given that there is an underlying BOLD signal, 
the particle filter is excellent at finding the most probable cause of 
the observed signal; however, given enough noise it will also determine
the most likely underlying signal; which may not be flat. Thus, it is 
important to investigate methods of determining false positives using 
average residuals.

\subsection{Signal with Low Noise}
%tests had noise of $ \{\sigma_y = .01, \sigma_x = .005\} $, where $\sigma_y$ is the
%measurement noise, and $\sigma_x$ is the wiener step size. Both signals used the
%parameters .
%The particle filter used the parameters defined in \autoref{tab:Prior} (\autoref{sec:PriorDistrib}),
%thus the particle filter was not centered over the correct values.

To begin the single voxel simulation; I generated a signal using the following parameters:
$\{\tau_0 = 1.45, \alpha = .3, E_0 = .47, V_0 = .044, \tau_s = 1.94, \tau_f = 1.99, \epsilon = 1.8\}$.
These same parameters will used again in the high noise case. The noise realizations were
then based on a measurement noise, $\sigma_y$, of $.001$ and a drift standard deviation,
$\sigma_x$ of $.0005$. The measurement noise as well as the steps of the drift
were taken to be Gaussian
For this low noise case, the eleven realizations are shown in \autoref{fig:LowNoiseRealization}.
\begin{figure}
\label{fig:LowNoiseRealization}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/realization_lownoise}
\caption{Test Signals with low noise compared to the clean signal}
The resulting fits compared to the \emph{true} BOLD signal are shown in \autoref{fig:FitComparisonLowNoise}.
While the error between the true signal and the estimated signal are low, the noise
is also relatively low. The actual signal that was delivered into the particle filter
is the result of some preprocessing to remove drift, as described in 
\autoref{sec:Preprocessing}. The preprocessed signal is compared
to the true BOLD signal in \autoref{fig:PreprocessedLowNoise}.
\begin{figure}
\label{fig:PreprocessedLowNoise}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/preprocessed_lownoise}
\caption{A comparison of the preprocessed signals for the low noise case. This is the
noisy input to the actual particle filter algorithm.}
The preprocessing consists of several steps, which are discussed in detail in \autoref{sec:Preprocessing}.
Obviously the spline struggles to match the signal near the end, as the graph shows, although 
overall the preprocessing seems to have been successful at removing trends. Finally, the 
set of fits to the preprocessed data, and in 
the true BOLD signal is shown in \autoref{fig:FitComparisonLowNoise}.
\begin{figure}
\label{fig:FitComparisonLowNoise}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/comparison_lownoise}
\caption{A comparison of the fitted signals for the low noise case.}
The filtering property of the particle filter is clear from the results here.  In several locations the output
of the particle filter looks like a filtered version of the input. For instance toward the
end, the estimates stay flat in spite of the preprocessed data drifting off a bit. By
this point, the algorithm had converged sufficiently to know that that such a movement in
signal isn't possible. A similar circumstance occurs at 100 seconds in. 
In general the fits are good, although because
of the slightly higher signal levels in the preprocessed signals, the fits tend to overestimate
the peaks. The final parameter sets are shown in \autoref{tab:LowNoiseParams}
\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c | c |}
\hline 
$\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$  & $ \sum \tau $ & $\sqrt{MSE}$ (Res.) &$\sqrt{MSE}$\\
\hline 
\rowcolor[gray]{.8}
1.45 & .3 & .47 & .044 & 1.94 & 1.99 & 1.8  & 5.38 &  & \\
\hline 
\hline 
1.22214 & 0.3449 & 0.33462 & 0.07138 & 1.60446 & 2.27529 & 1.59454 & 5.10189  &  0.00321067  & 0.00987647  \\
1.37493 & 0.33183 & 0.36296 & 0.07327 & 1.64076 & 2.10299 & 1.57627 & 5.11867 &  0.00305475  & 0.00993227  \\
1.16604 & 0.32205 & 0.34057 & 0.08217 & 1.64768 & 2.35351 & 1.24515 & 5.16722 &  0.00328932  & 0.00967961  \\
1.23176 & 0.32707 & 0.34028 & 0.07961 & 1.62698 & 2.18519 & 1.30326 & 5.04394 &  0.00284719  & 0.00912005  \\
1.1832 & 0.31787 & 0.34718 & 0.0821 & 1.54961 & 2.29115 & 1.27817 & 5.02396   &  0.00300634  & 0.00971258  \\
1.1424 & 0.33395 & 0.34725 & 0.07366 & 1.62208 & 2.29084 & 1.4025 & 5.05531   &  0.00283287  & 0.00948483  \\
1.30041 & 0.3596 & 0.35643 & 0.07679 & 1.56406 & 2.13234 & 1.60338 & 4.99681  &  0.00302802  & 0.01021885  \\
1.24008 & 0.34601 & 0.33978 & 0.08903 & 1.64989 & 2.23655 & 1.29004 & 5.12651 &  0.00304378  & 0.01007964  \\
1.1709 & 0.32739 & 0.34644 & 0.08255 & 1.53734 & 2.28262 & 1.37828 & 4.99087  &  0.00334488  & 0.01032886  \\
1.18967 & 0.34344 & 0.33554 & 0.07976 & 1.53582 & 2.30746 & 1.42774 & 5.03295 &  0.00317542  & 0.01001503  \\
1.184 & 0.34053 & 0.35017 & 0.08917 & 1.61025 & 2.27926 & 1.16448 & 5.07352   &  0.00288855  & 0.00950536  \\
\hline                                                                           
1.21868 & 0.33588 & 0.34557 & 0.07995 & 1.59899 & 2.24884 & 1.38762 & 5.06651 & 0.00306562     & 0.00981396 \\
\hline 
\end{tabular}
\caption{Estimated Parameters on 11 different runs with low noise. First row is the true parameters,
last is mean over the 10 runs. The $\sqrt{MSE}$ (Res.) is the square root of the mean square of the
residuals. Essentially this is the is the MSE between the estimated signal and the noisy signal which 
was available to the particle filter. Square root of MSE is the actual \emph{error}, with the true signal,
which of course was not available to the particle filter.}
\label{tab:LowNoiseResults} 
\end{table}
There are a few results worth noting here. First the time constants vary greatly across
runs with different noise realizations, yet the sum of the individual time constants
($\tau_f$, $\tau_s$ and $\tau_0$) seem to be more consistent. In general the 
time constants are falling short of the true time constant. This could be a limitation
based on the prior distribution (which notably has an initial mean below the values used
in the simulation) or it could be caused by the dissassociated benefits of a correct time
constant with correct output. It often takes several measurement periods before a difference
in time constants becomes relevent. It is possible that in that time other parameters 
may be more relevent. As a result, by the time the time constant has made a substantial
difference in the weight, the 
particle with the better $\tau$ will have been weighted based on other parameters several
times. Another interesting result in the huge variation in the levels of $V_0$. In general,
with this admittedly small amount of noise, it would appear that the relation between a set
of parameters/stimuli and the output is not injective. In other words, it would appear
that a timeseries is not unique to a single set of parameters. This is good justification that 
simultaneous blood volume or tagged flow calculations with the conventional FMRI 
could definitely benefit the model. 
\begin{table}[t]
\centering
\begin{tabular}{|c | c  c  c  c  c  c  c |}
\hline 
  & $\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$ \\
\hline 
\rowcolor[gray]{.8} $\tau_0$ &   0.0189481 & -0.0014269 & -0.0011267 & -1.13e-05 & -0.0025616 & -0.0189559 & 0.0070405 \\
$\alpha$ &                       -0.0014269 & 0.0026716 & 9.93e-05 & -0.0002041 & -0.0008632 & -0.0016823 & 0.0071891 \\
\rowcolor[gray]{.8} $E_0$    &   -0.0011267 & 9.93e-05 & 0.0010701 & -0.0002277 & -0.0001177 & 0.0001013 & 0.0016972 \\
$V_0$    &                       -1.13e-05 & -0.0002041 & -0.0002277 & 0.0005401 & 4.3e-06 & 4.56e-05 & -0.0080494 \\
\rowcolor[gray]{.8} $\tau_s$ &   -0.0025616 & -0.0008632 & -0.0001177 & 4.3e-06 & 0.0128056 & 0.012878 & -0.005516 \\
$\tau_f$ &                       -0.0189559 & -0.0016823 & 0.0001013 & 4.56e-05 & 0.012878 & 0.0416927 & -0.0158182 \\
\rowcolor[gray]{.8} $\epsilon$&  0.0070405 & 0.0071891 & 0.0016972 & -0.0080494 & -0.005516 & -0.0158182 & 0.1567165 \\
\hline 
\end{tabular}
\caption{Typical Covariance matrix of the parameters at the end of a run.}
\label{tab:CovSim} 
\end{table}

@@

\begin{figure}
\subfigure[Converging histogram for $Q$, $S$, $F$, and $BOLD$ of the first run, low noise simulation.]
{\includegraphics[trim=7cm 2cm 7cm 2cm, width=15cm]{images/converge_lownoise3}}
\label{fig:LowNoiseHist}
\end{figure}
The covariance matrix (\autoref{tab:CovSim}) also confirms the idea that the $\tau$ parameters are interchangeable
as far as the BOLD signal goes. Notice the covariance of $\tau_f$ and $\tau_0$ is $-0.019$ whereas
the variance of $\tau_0$ and $\tau_f$ are $0.019$ and $0.04$ respectively. Clearly there is 
some measure of ill-defined behavior between the differen time-constants, as well as several
other variables.  The convergence properties of the first run in \autoref{tab:LowNoiseResults} may be
viewed in \autoref{fig:LowNoiseHist}.

TODO more about convergence
